---
title: "How has human mobility changed under the COVID-19 Pandemic?"
author: "Kelsey Gonzalez"
date: "April 15, 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

This visualization aims to understand population behavior during the COVID-19 pandemic. I use aggregated human mobility data (how far people move each day) from Cuebiq, an offline intelligence and measurement marketing company, which partners with 86 apps to collect first-party location data. Different U.S. states have handled the Covid-19 Pandemic in vastly different ways, with some states still refraining from issuing stay-at-home orders. This has led to very different mobility behaviors by state. However, this visualization shows that the stay-at-home order successfully reduced movement during the week of March 23rd, with Americans quickly reverting towards pre-COVID-19 movement patterns. 

Data is available at https://www.cuebiq.com/visitation-insights-covid19/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(gganimate)
library(ggthemes)
library(choroplethr)
library(maps)
library(htmltools)
require(transformr)

theme_set(theme_minimal())

theme_map <- function(...) {
  theme_minimal() +
    theme(
      axis.line = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      plot.background = element_rect(fill = "white", color = NA), 
      panel.background = element_rect(fill = "white", color = NA), 
      legend.background = element_rect(fill = "white", color = NA),
      legend.title = element_blank(), 
      panel.border = element_blank(),
      legend.position = "none",
      plot.caption = element_text(hjust = 1, face= "italic"),
      plot.title.position = "plot",
      plot.caption.position =  "plot"
    )
}

```

```{r data wrangling, include=FALSE}
covid <- read_csv("covid.csv") %>% 
  pivot_wider(id_cols = c(state, state_stub), 
              names_from = week_date_mdy, 
              values_from = mobility)  %>% 
  mutate( `2020-03-02` = (`2020-03-02`) / `2020-02-24`,
          `2020-03-09` = (`2020-03-09`) / `2020-02-24`, 
          `2020-03-16` = (`2020-03-16`) / `2020-02-24`, 
          `2020-03-23` = (`2020-03-23`) / `2020-02-24`, 
          `2020-03-30` = (`2020-03-30`) / `2020-02-24`, 
          `2020-04-06` = (`2020-04-06`) / `2020-02-24`,
          `2020-04-13` = (`2020-04-13`) / `2020-02-24`,
          `2020-02-24` = 1) %>% 
  pivot_longer(cols = -c(state, state_stub), names_to = "date", values_to = "mobility", names_prefix = "mobility_") %>% 
  mutate(date = lubridate::ymd(date)) 

```

```{r plot, include=FALSE}
interesting <- c(
  "Alabama",  "South Carolina", "Georgia",
  "New York",
  "California", "Washington", "Arizona", "New Jersey"
)

covid <-
  covid %>% mutate(
    colors = ifelse(state %in% interesting, "interesting", "not"),
    lab = ifelse(state %in% interesting, str_to_upper(state_stub), "")
  )

a <- ggplot(covid, aes(x = date,
                       y = mobility,
                       group = state_stub)) +
  geom_line(aes(color = colors),
            show.legend = FALSE,
            alpha = 0.7) +
  scale_color_manual(values = c('black', 'grey80')) +
  geom_point(
    data = covid %>% filter(colors != "interesting"),
    shape = 21,
    colour = 'grey80',
    fill = 'grey80',
    alpha = 0.7,
    size = 2
  ) +
  geom_point(
    data = covid %>% filter(colors == "interesting"),
    aes(fill = mobility),
    shape = 21,
    colour = "black",
    alpha = 0.7,
    size = 4,
    stroke = 2
  ) +
  scale_fill_gradient(low = "forestgreen", high = "white") +
  geom_text(
    aes(label = lab),
    show.legend = FALSE,
    hjust = -.5,
    vjust = .3
  ) +
  xlab("") +
  ylab("Mobility Rate") +
  scale_x_date(date_breaks = "1 week",
               date_labels = "%m-%d") +
  scale_y_continuous(limits = c(.2, 1.1)) +
  transition_reveal(along = date) +
  ggtitle("How has human mobility changed under the COVID-19 pandemic?",
          subtitle = "Week of: {frame_along}") +
  view_follow(fixed_y = T)

animate(a, height = 400, end_pause = 10)
anim_save("output-gif/plot.gif")
```

```{r map, include=FALSE}
# code influenced by https://ditheringdata.netlify.app/2018/01/01/gganimate/

us <- map_data("state") %>% rename(state = region) %>% mutate(state = str_to_title(state))
mobility <- covid %>% left_join(us)

b <- ggplot(data = mobility,
            aes(frame = date)) +
  geom_polygon(aes(
    x = long,
    y = lat,
    group = group,
    fill = mobility
  ),
  color = "#8c8c8c") +
  theme_map() +
  scale_fill_gradient(low = "forestgreen",
                      high = "white") +
  coord_map("albers",
            lat0 = 30,
            lat1 = 40) +
  labs(caption = "Mobility data from Cuebiq\nChart by @kelseyegonzalez") +
  transition_states(
    states = date,
    transition_length = 1.5,
    state_length = 1
  )

animate(b, height = 400, end_pause = 10)
anim_save("output-gif/map.gif")
```


```{r combine gifs}
# code to combine gifs from https://github.com/thomasp85/gganimate/wiki/Animation-Composition

a_mgif <- magick::image_read("output-gif/plot.gif")
b_mgif <- magick::image_read("output-gif/map.gif")

new_gif <- magick::image_append(c(a_mgif[1], b_mgif[1]))
for (i in 2:100) {
  combined <- magick::image_append(c(a_mgif[i], b_mgif[i]))
  new_gif <- c(new_gif, combined)
}

new_gif
# anim_save("together.gif")
```

Source code can be found https://github.com/kelseygonzalez/covid_mobility
